---
title: "650 Report R file"
output: html_notebook
---

# Import data set 
imports the 2 data sets that finds the common genes in the two data sets. Forming data table only using the intersect of genes (features) and rbinding the two data sets. 
```{r}
library(dplyr)

Pheno <- read.delim("Pheno.txt")
Expr <- read.delim("Expr.txt")

Pheno1 <- read.delim("Pheno(1).txt")
Expr1 <- read.delim("Expr(1).txt")

Expr <- t(Expr)
dt1 <- cbind.data.frame(Expr, Progression = Pheno$Progression)

Expr1 <- t(Expr1)
dt2 <- cbind.data.frame(Expr1, Progression = Pheno1$Progression)

cols <- intersect(colnames(dt1), colnames(dt2))
which('Progression' == cols) == length(cols)

dt <- rbind.data.frame(dt1[,cols], dt2[,cols])
dim(dt)

```

separate progression sampels from non-progression samples
```{r}
dt_prog <- dt%>%filter(Progression == 'Progression')
dt_non_prog <- dt%>%filter(Progression == 'NoProgression')

ps <- sort(sample(nrow(dt_prog), nrow(dt_prog)*.7))
ns <- sort(sample(nrow(dt_non_prog), nrow(dt_non_prog)*.7))

test <- rbind.data.frame(dt_prog[-ps,],dt_non_prog[-ns,])
train <- rbind.data.frame(dt_prog[ps,],dt_non_prog[ns,])

dt <- train
dt_prog <- dt%>%filter(Progression == 'Progression')
dt_non_prog <- dt%>%filter(Progression == 'NoProgression')

```

# Explore Key genes using Wilcoxon test
Generate Wilcoxon test table
```{r}
WilcoxPvalTable <- data.frame(NULL)
for(i in 1:(ncol(dt)-1)){
  p <- dt_prog[,i]
  np <- dt_non_prog[,i]
  name <- colnames(dt)[i]
  wx <- wilcox.test(p,np)
  pval <- wx$p.value
  add <- data.frame(gene = name, pval = pval)
  WilcoxPvalTable <- rbind.data.frame(WilcoxPvalTable, add)
}

hist(WilcoxPvalTable$pval, breaks = 50, xlab = 'p-Values', main = 'Histogram of Wilcoxon test p-values', col = rgb(0,0,1,1/4))

WilcoxPvalTable <- WilcoxPvalTable%>%arrange(pval)
```


Inspect expression level histogram
```{r}
test_col <- 'PSRC1'
wilcox.test(dt_prog[,test_col], dt_non_prog[,test_col])
p1 <- hist(dt_non_prog[,test_col])
p2 <- hist(dt_prog[,test_col])
plot( p1, col=rgb(0,0,1,1/4), main = test_col, xlab = 'Expression Level')  
plot( p2, col=rgb(1,0,0,1/4), add=T)
legend("topright", legend=c("non-Progression", "Progression"),
       col=c(rgb(0,0,1,1/4), rgb(1,0,0,1/4)), fill=c(rgb(0,0,1,1/4), rgb(1,0,0,1/4)), cex=0.8)
```

# Explore Key genes with GWAS
change to categorical data, over/normal/under expressed decided by Q13 values of non-progression set
```{r}
dt_cat <- dt
for(i in 1:(ncol(dt)-1)){
  Q <- quantile(dt_non_prog[,i])
  dt_cat[dt[,i] < Q[2],i] <- 'U'
  dt_cat[dt[,i] >= Q[2],i] <- 'N'
  dt_cat[dt[,i] > Q[4],i] <- 'O'
}

dt_prog_cat <- dt_cat%>%filter(Progression == 'Progression')
dt_non_prog_cat <- dt_cat%>%filter(Progression == 'NoProgression')
```


perform GWAS of categorized data
```{r}
Gt_prog <- data.frame(NULL)
for(i in 1:(ncol(dt)-1)){
  add <- table(dt_prog_cat[,i])
  Gt_prog <- rbind.data.frame(Gt_prog, add)
}
rownames(Gt_prog) <- colnames(dt)[1:(ncol(dt)-1)]
colnames(Gt_prog) <- c('N','O','U')

Gt_non_prog <- data.frame(NULL)
for(i in 1:(ncol(dt)-1)){
  add <- table(dt_non_prog_cat[,i])
  Gt_non_prog <- rbind.data.frame(Gt_non_prog, add)
}
rownames(Gt_non_prog) <- colnames(dt)[1:(ncol(dt)-1)]
colnames(Gt_non_prog) <- c('N','O','U')

GWAS <- data.frame(NULL)
for(i in 1:(ncol(dt)-1)){
  for(j in c('N','O','U')){
    freq_p <- Gt_prog[i,j]
    freq_n <- Gt_non_prog[i,j]
    add <- data.frame(prog = freq_p, nonprog = freq_n)
    rownames(add) <- paste(j, '_', colnames(dt)[i], sep = "")
    GWAS <- rbind.data.frame(GWAS, add)
  }
}

GWAS <- t(GWAS)
dim(GWAS)

GWAS_pval <- GWAS
options(warn = -1)
for(j in 1:ncol(GWAS)){
  r <- which(rownames(GWAS) == rownames(GWAS)[1])
  c <- which(colnames(GWAS) == colnames(GWAS)[j])
  caseF <- GWAS[r,c]/sum(GWAS[-r,c])
  controlF <- sum(GWAS[r,-c])/sum(GWAS[-r,-c])
  odds <- caseF/controlF
  chiPval <- chisq.test(c(odds,1))
  GWAS_pval[1,j] <- chiPval$p.value
}

Glp <- -log(GWAS_pval[1,])
plot(Glp, col = 'blue', pch = 16, main = 'Manhattan plot', ylab = '-logPval')
hist(GWAS_pval[1,], breaks = 50, xlab = 'p-Values', main = 'Histogram of GWAS p-values', col =rgb(1,0,0,1/4))
```

Compare Wilcoxon and GWAS selection results
```{r}
Glp <- sort(Glp, decreasing = TRUE)

rank_sum <- 0
top100 <- c()
for(i in 1:100){
  gene <- names(Glp)[i]
  gene <- strsplit(gene, '_')[[1]][2]
  rank <- which(gene == WilcoxPvalTable$gene)
  rank_sum <- rank_sum + rank
  top100 <- c(top100, gene)
}
rank_sum/100
```

Inspect GWAS results
```{r}
test_col <- 'ALOX5'#strsplit(names(Glp)[1], '_')[[1]][2]
wilcox.test(dt_prog[,test_col], dt_non_prog[,test_col])
p1 <- hist(dt_non_prog[,test_col])
p2 <- hist(dt_prog[,test_col])
plot( p1, col=rgb(0,0,1,1/4), main = colnames(dt)[test_col], xlab = 'Expression Level')  
plot( p2, col=rgb(1,0,0,1/4), add=T)

t.test(dt_non_prog[,test_col],dt_prog[,test_col])
#legend("topright", legend=c("non-Progression", "Progression"),
#       col=c(rgb(0,0,1,1/4), rgb(1,0,0,1/4)), fill=c(rgb(0,0,1,1/4), rgb(1,0,0,1/4)), cex=0.8)
```

# Random forest training
Random forest training with GWAS top100 features
```{r}
train_cat <- train
for(i in 1:(ncol(dt)-1)){
  Q <- quantile(dt_non_prog[,i])
  train_cat[train[,i] < Q[2],i] <- 'U'
  train_cat[train[,i] >= Q[2],i] <- 'N'
  train_cat[train[,i] > Q[4],i] <- 'O'
}

test_cat <- test
for(i in 1:(ncol(dt)-1)){
  Q <- quantile(dt_non_prog[,i])
  test_cat[test[,i] < Q[2],i] <- 'U'
  test_cat[test[,i] >= Q[2],i] <- 'N'
  test_cat[test[,i] > Q[4],i] <- 'O'
}

dt_cat <- train_cat[,c(top100, 'Progression')]
dt_cat$Progression <- as.factor(dt_cat$Progression)
test_cat <- test_cat[,c(top100, 'Progression')]
test_cat$Progression <- as.factor(test_cat$Progression)

rf <- randomForest::randomForest(Progression~., data=dt_cat)
```

draw roc curve 
```{r}
predictions <- as.data.frame(predict(rf, test_cat, type = "prob"))
predictions <- predictions$Progression
label <- as.numeric(test_cat$Progression)-1
precrec_obj <- precrec::evalmod(scores = predictions, labels = label)
plot(precrec_obj)
t1 <- which(precrec_obj$rocs[[1]]$y > 0.8)
t2 <- which(precrec_obj$rocs[[1]]$y < 0.9)
t <- intersect(t1,t2)
spec80 <- 1 - min(precrec_obj$rocs[[1]]$x[t])
```
Inspect features
```{r}
gene_imp <- as.data.frame(rf$importance)
gene_imp <- gene_imp%>%arrange(desc(MeanDecreaseGini))
```

# Nerual network training
TrainTest processing
```{r}
nn_train <- train[,c(top100, 'Progression')]
nn_test <- test[,c(top100, 'Progression')]
nn_train$Progression <- as.factor(nn_train$Progression)
nn_test$Progression <- as.factor(nn_test$Progression)
nn_train$Progression <- as.numeric(nn_train$Progression)-1
nn_test$Progression <- as.numeric(nn_test$Progression)-1
```

4-fold CV for architecture
```{r}
nn_train_prog <- nn_train%>%filter(Progression == 1)
nn_train_non_prog <- nn_train%>%filter(Progression == 0)


arch_search <- data.frame(NULL)
for(i in 1:3){
  for(j in 1:100){
  sum1 <- 0
  sum2 <- 0
    for(k in 1:20){
      ps <- sort(sample(nrow(nn_train_prog), nrow(nn_train_prog)*.75))
      ns <- sort(sample(nrow(nn_train_non_prog), nrow(nn_train_non_prog)*.75))
      
      nn_test_cv <- rbind.data.frame(nn_train_prog[-ps,],nn_train_non_prog[-ns,])
      nn_train_cv <- rbind.data.frame(nn_train_prog[ps,],nn_train_non_prog[ns,])
      
      nn <- neuralnet::neuralnet(Progression~.,data=nn_train_cv, hidden=rep(j,i), act.fct = "logistic",
                      linear.output = FALSE)
      
      Predict <- neuralnet::compute(nn,nn_test_cv)
      prob <- Predict$net.result
      precrec_obj <- precrec::evalmod(scores = prob, labels = nn_test_cv$Progression)
      t1 <- which(precrec_obj$rocs[[1]]$y > 0.8)
      t2 <- which(precrec_obj$rocs[[1]]$y < 0.9)
      t <- intersect(t1,t2)
      spec80 <- 1 - mean(precrec_obj$rocs[[1]]$x[t])
      auc <- attr(precrec_obj$rocs[[1]],'auc')
      sum1 <- auc + sum1
      sum2 <- spec80 + sum2
    }
  auc <- sum1/20
  spec <- sum2/20
  add <- data.frame(layer = i, nodes = j, spec80 = spec, auc = auc)
  arch_search <- rbind.data.frame(arch_search, add)
  print(paste(j, ' of ',i ,sep = ''))
  }
}


```

Construct optimum NN
```{r}
arch_search <- arch_search%>%arrange(desc(spec80))
nn <- neuralnet::neuralnet(Progression~.,data=nn_train, hidden=c(60,60,30), act.fct = "logistic",
                linear.output = FALSE)

Predict <- neuralnet::compute(nn,nn_test)
prob <- Predict$net.result
precrec_obj <- precrec::evalmod(scores = prob, labels = nn_test$Progression)
t1 <- which(precrec_obj$rocs[[1]]$y > 0.8)
t2 <- which(precrec_obj$rocs[[1]]$y < 0.9)
t <- intersect(t1,t2)
spec80 <- 1 - min(precrec_obj$rocs[[1]]$x[t])
auc <- attr(precrec_obj$rocs[[1]],'auc')
plot(precrec_obj)
spec80
auc

```


